Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-03-03 doc/paper/paper.tex
First draft of paper! Yesgit add paper.pdf -f

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{GTR}{GTR}{generalized time-reversible}

\graphicspath{{figures/}}

@@

The structure of the contact network underlying a viral epidemic can profoundly
impact the speed and pattern of the epidemic's spread. Network structure can
influence the prevalence trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}. True contact networks are a challenging type
of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that statistical
inference might be possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
to as ``scale-free''. The first contact network model incorporating
preferential attachment was introduced by \textcite{barabasi1999emergence}, and
is now referred to as the \gls{BA} model. Under this model, networks are formed
by iteratively adding nodes with $m$ new edges each. These new edges are joined
to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} fit a preferential attachment
model to a phylogenetically estimated transmission network relating 60\% of HIV
infected \gls{MSM} in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection.

Due to their complexity, it is generally difficult to explicitly calculate the
likelihood of a transmission tree under a contact network model. To do so, we
would need to integrate over all possible networks generated by that model, and
also over all possible labellings of the internal nodes of the transmission
tree. While it is not known (to us) whether such integration is tractable, a
simpler alternative is offered by likelihood-free methods, namely
\gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring}. \gls{ABC}
leverages the fact that, although calculating the likelihood may be impossible,
generating simulated datasets according to a model is often straightforward. If
our model fits the data well, the simulated data it produces should be similar
to the observed data. More formally, if $D$ is the observed data, the posterior
distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
dataset according to $\theta$, and $\varepsilon$ is a small
tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
is a kernel function, the approach is known as
kernel-\gls{ABC}~\autocite{nakagome2013kernel}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree, using the tree
kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
of the \gls{BA} model on a variety of simulated and real data sets.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).

@@

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. The parameters were varied one at a time over the
following values: $N$ = \sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000},
$m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.5, 1, 1.5}. For each parameter
set, 100 networks were generated, and a transmission tree was simulated over
each (300 trees per parameter set). The 300 trees were compared pairwise with
the tree kernel to form a $300 \times 300$ kernel matrix. We constructed a
\gls{kSVM} classifier for the parameter of interest using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations. The construction of the kernel
matrices and classifiers was repeated for several combinations of the kernel
meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
basis function variance'')~\autocite[see][]{poon2013mapping}, and for
transmission trees of sizes 100, 500, and 1000. We also tested univariate
classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To ensure our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method \autocite[originally
used by][]{sisson2007sequential}. We were able to obtain a close approximation
to the function (see Figure~\ref{fig:smctest}), and attained the stopping
condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1, 1.5}.
The adaptive \gls{ABC} algorithm was applied to each tree with these priors:
$m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
uniform on the triangular region \{$500 \leq N \leq 15000$, $500 \leq I \leq
 5000$, $I \leq N$\}. Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. The algorithm was run
with 1000 particles, 5 simulated datasets per particle, and the ``quality''
parameter set to 0.95. We use the same stopping criterion as
\textcite{del2012adaptive}, namely when the MCMC acceptance rate dropped below
1.5\%. Point estimates for the parameters were obtained by taking the highest
point of an estimated kernel density on the final set of particles.

@@

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. Characteristics of the datasets we
investigated are given in Table~\ref{tab:data}.

\begin{table}
  \centering
  \begin{tabular}{ccccc}
    Reference & Sequences ($n$) & Location & Risk group & Gene \\
    \hline
    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{1299} & \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} & \multirow{2}{*}{\textit{env}} \\
    \autocite{novitsky2014impact} \\
    \hline
  \end{tabular}
  \caption{Characteristics of published datasets investigated with kernel-ABC.
  Acronyms: MSM, men who have sex with men.}
  \label{tab:data}
\end{table}

We downloaded all sequences associated with each study from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31
\autocite{edgar2004muscle}, and alignments were manually inspected with
\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
\gls{GTR} model. Transmission trees were estimated by rooting and time-scaling
the phylogenies by root-to-tip regression, using a modified version of
Path-o-Gen \autocite[distributed as part of \software{BEAST},
][]{drummond2007beast}, as described
previously~\autocite{poon2015phylodynamic}.

Two additional steps were undertaken for the
\textcite{novitsky2013phylogenetic, novitsky2014impact} data. First, before
performing the multiple sequence alignment, we aligned each
sequence pairwise to the HXB2 reference sequence (GenBank accession number
HIVHXB2CG) using \software{MAFFT} version 7.130b \autocite{katoh2013mafft} and
clipped out the hypervariable regions with \software{Biopython}
version 1.66+ \autocite{cock2009biopython}. Second, after multiple sequence
alignment, the sequences were trimmed to an informative region with
\software{trimAl} version 1.4.rev15 \autocite{capella2009trimal}.

Kernel-ABC was used to fit the \gls{BA} model to each dataset. To make the
analyses comparable, the same priors and configuration parameters were used for
each dataset. The priors were as follows: $\alpha \sim$ Uniform(0, 2), $m \sim$
Uniform(1, 5), and $(N, I) \sim$ Uniform(\sett{$n \leq N \leq 10000$, $n \leq I
\leq 1000$, $I \leq N$}), where $n$ is the number of sequences in the dataset.
The kernel meta-parameters and SMC configuration were the same as in the
\gls{ABC} simulations.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 
    0.8 
for all combinations of tree size and epidemic size. The $I$ parameter was also
classified accurately, with an average $R^2$ of 
    0.93 
for 500-tip trees or 
    0.7 
for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
varying between 
    0.004
and
    0.36
depending on $I$ and the tree size. Finally, the accuracy of the classifier for
$N$ varied widely, from 
    0.08
to
    0.82.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

\begin{figure}
  \includegraphics{kernel-alpha-crossv}
  \caption{Cross-validation $R^2$ of kernel-SVM classifier for the $\alpha$
    parameter of the \gls{BA} network model.}
  \label{fig:crossv}
\end{figure}

@@

Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
The results for the other values of $m$ are similar and can be found in the
supplemental materials.

The marginal approximated posterior distributions for one simulated dataset are
shown in Figure~\ref{fig:abcex}.

\begin{figure}
    \includegraphics{abc-point-est}
    \caption{Point estimates of \gls{BA} model parameters obtained by running
    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
    and limits of the $y$-axes are the regions of uniform prior density.}
    \label{fig:abcpt}
\end{figure}

\begin{figure}
    \includegraphics{abc-example}
    \caption{Marginal estimated posterior distributions obtained with ABC on a
    single simulated dataset. Dotted lines and shaded polygon indicate true
    values.}
    \label{fig:abcex}
\end{figure}

@@

% kernel classifiers

The $m$ and $N$ parameters were much harder to recover than $\alpha$ and $I$,
suggesting differing influences of these parameters on tree shape. The $\alpha$
parameter in particular was accurately estimated in most cases, implying that
the strength of preferential attachment has a strong impact on the shape of the
transmission tree.

% abc simulations

>>>>

\usepackage{alltt}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{multirow}
\usepackage{cleveref}

@@

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}
\newacronym{HPD}{HPD}{highest posterior density}
\newacronym{GTR}{GTR}{general time-reversible}
\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
\newacronym{IDU}{IDU}{injection drug user}

\graphicspath{{../figures/}}
\newcommand{\tablepath}{../tables}
\def\figwidth{0.8\textwidth}

@@

When an infectious disease spreads through a population, transmissions cannot
generally occur between any two individuals at any time. Rather, the
individuals must have a particular mode of contact with one another, which
varies with the disease's mode of transmission. For airborne pathogens,
physical proximity may be sufficient, but for sexually transmitted diseases,
either sexual or blood-to-blood contact is required. The set of links between
individuals along which transmission can occur is called the contact
network~\autocite{klovdahl1985social}. The structure of the contact network
underlying an epidemic can profoundly impact the speed and pattern of the
epidemic's expansion. Network structure can influence the prevalence
trajectory~\autocite{o2010contact} and epidemic
threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
of quantities such as effective population
size~\autocite{goodreau2006assessing}. From a public health perspective,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{wang2015targeting}. True contact networks are a challenging
type of data to collect, requiring extensive epidemiological
investigation~\autocite{welch2011statistical}.

Viral sequence data, on the other hand, has become relatively inexpensive and
straightforward to collect on a population level. Due to the high mutation rate
of RNA viruses, epidemiological processes impact the course of viral evolution,
thereby shaping the intra-host viral
phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
coined to describe this interaction, as well as the growing family of inference
methods to estimate epidemiological parameters from viral
phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
diverse properties of local viral outbreaks, from basic reproductive
number~\autocite{stadler2011estimating}, to the degree of
clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
during acute infection~\autocite{volz2012simple}. On the other hand, inference
of structural network parameters has to date been limited. However, it has been
shown that network structure has a tangible impact on phylogeny
shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
inference might be possible~\autocite{welch2011statistical}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[although there has been
some disagreement, see][]{jones2003assessment}. Under these models, nodes with
a high number of contacts attract new connections at an elevated rate. Networks
produced by preferential attachment have a power-law degree distribution,
meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
for some constant $\gamma$. These networks are also referred to as
``scale-free''. The first contact network model incorporating preferential
attachment was introduced by \textcite{barabasi1999emergence}, and is now
referred to as the \gls{BA} model. Under this model, networks are formed by
iteratively adding nodes with $m$ new edges each. These new edges are joined to
existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
that nodes of high degree tend to attract more connections (in the original
paper, only $\alpha = 1$ was investigated).

Previous work offers precedent for the possibility of statistical inference of
structural network parameters. \textcite{britton2002bayesian} develop a
Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
network~\autocite{erdos1960evolution} given observed infection dates, and
optionally recovery dates. Their approach was later extended by
\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
individuals. \textcite{brown2011transmission} found that a Waring distribution,
which is produced by a different preferential attachment model, was a good fit
to a phylogenetically estimated transmission network relating 60\% of HIV
infected \gls{MSM} in the United Kingdom. The transmission network is a
subgraph of the contact network which includes only those edges which have
already led to a new infection.

Standard methods of model fitting involve calculation of the likelihood of
observed data under the model. In maximum likelihood estimation, a quantity
proportional to the likelihood is optimized, often through a standard
multi-dimensional numerical optimization procedure. Bayesian methods integrate
prior information by optimizing the posterior probability instead. To avoid
calculation of a normalizing constant, Bayesian inference is often performed
using \gls{MCMC}, which requires the calculation of likelihood ratios.
Unfortunately, it is generally difficult to explicitly calculate the likelihood
of an observed transmission tree under a contact network model. To do so, it
would be necessary to integrate over all possible networks, and also over all
possible labellings of the internal nodes of the transmission tree. While it is
not known (to us) whether such integration is tractable, a simpler alternative
is offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages
the fact that, although calculating the likelihood may be impossible,
generating simulated datasets according to a model is often straightforward. If
our model fits the data well, the simulated data it produces should be similar
to the observed data. More formally, if $D$ is the observed data, the posterior
distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
dataset according to $\theta$, and $\varepsilon$ is a small
tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
is a kernel function, the approach is known as kernel-\gls{ABC}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree, using the tree
kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
of the \gls{BA} model on a variety of simulated and real data sets. Our results
show that the attachment power parameter $\alpha$ can be inferred with
reasonable accuracy, and can vary considerably between epidemics from different
settings.

@@

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
in the same fashion as several previous studies
\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
that our implementation was correct, we reproduced Figure 1A of
\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
unbalancedness of transmission trees simulated over four network models at
various levels of pathogen transmissibility.

@@

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. The
parameters were varied one at a time over the following values: $N$ =
\sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000}, $m$ = \sett{2, 3, 4},
and $\alpha$ = \sett{0.5, 1, 1.5}. The parameters not being tested were fixed
at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ = 1. For each
parameter set, 100 networks were generated, and a transmission tree was
simulated over each (300 trees per parameter set). The 300 trees were compared
pairwise with the tree kernel to form a $300 \times 300$ kernel matrix. We
constructed a \gls{kSVM} classifier for the parameter of interest using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations. The construction of the kernel
matrices and classifiers was repeated for several combinations of the kernel
meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
basis function variance'')~\autocite[see][]{poon2013mapping}, and for
transmission trees of sizes 100, 500, and 1000. We also tested univariate
classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To check that our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
1.5}. This produced a total of 24 parameter combinations $\times$ three trees
per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
were Gaussian, with variance equal to twice the empirical variance of the
particles. The algorithm was run with 1000 particles, 5 simulated datasets per
particle, and the ``quality'' parameter controlling the decay rate of the
tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
dropped below 1.5\%. Point estimates for the parameters were obtained by taking
the highest point of an estimated kernel density on the final set of particles,
using the \software{density} function with the default parameters in
\software{R}. \gls{HPD} intervals were calculated with the
\software{HPDinterval} function from the \software{R} package
\software{coda}~\autocite{plummer2006coda}.

@@

We applied our kernel-ABC method to several published HIV datasets. Because the
\gls{BA} model generates networks with a single connected component, we
specifically searched for datasets which originated from existing clusters,
either phylogenetically or geographically defined. Characteristics of the
datasets we investigated are given in \cref{tab:data}.

\begin{table}[ht]
  \centering
  \input{\tablepath/realdata}
  \caption{Characteristics of published datasets investigated with kernel-ABC.
  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
  \label{tab:data}
\end{table}

We downloaded all sequences associated with each study from GenBank. For the
\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
hypervariable regions were clipped out with BioPython version
1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
were manually inspected with \software{Seaview} version 4.4.2
\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
alignments by approximate maximum likelihood using \software{FastTree2} version
2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
and time-scaling the phylogenies by root-to-tip regression, using a modified
version of Path-O-Gen (distributed as part of
BEAST~\autocite{drummond2007beast}) as described
previously~\autocite{poon2015phylodynamic}.

Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
larger than the others, containing 1265 and 1299 sequences respectively. To
ensure that the analyses were comparable, we reduced these to a number of
sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
described previously~\autocite{poon2014impact}. Only sequences within this
cluster were carried forward. For the \textcite{novitsky2014impact} data, no
large clusters were detected using the same cutoff, so we analysed a subtree of
size 180 chosen arbitrarily.

@@

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
tree kernel, although the magnitude of the disparity varied between the
parameters (\cref{fig:crossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For
appropriate choices of the kernel meta-parameters, the kSVM classifier for
$\alpha$ had an $R^2$ above 
    0.8 
for all combinations of tree size and epidemic size (\cref{fig:crossv}). The
other classifiers' $R^2$ averaged between
  0.49
and
  0.61
for the SVM based on the \gls{nLTT}, and between
  0.73
and
  0.76
for the linear regression against Sackin's index. The $I$ parameter was also
classified accurately by the tree kernel (\cref{fig:Icrossv}), with an average
$R^2$ of 
  0.93 
for 500-tip trees or 
  0.7 
for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
the tree kernel, with $R^2$ values of
  0.83
for 500-tip trees or
  0.55
for 100-tip trees. Sackin's index was an extremely poor classifier for this
parameter, with both $R^2$ values below
  0.08.

The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
for the kSVM varying between 
  0.004
and
  0.36
depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
with $R^2$ below
  0.02
for all scenarios. Sackin's index fared slightly better, especially on trees
with a higher number of tips, with $R^2$ between
  0
and
  0.18.
Finally, the accuracy of the kSVM classifier for $N$ varied widely
(\cref{fig:Ncrossv}), from 
  0.08
to
  0.82.
The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
500-tip transmission trees ($R^2$ = 
  0.31
and
  0.53),
although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
  0.07).
Sackin's index could not classify $N$, with all $R^2$ below
  0.11.
Based on inspection of the cross-validation results, we chose to use the
meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
to use Sackin's index or the \gls{nLTT}.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{kernel-rsquared}
  \caption{
    Cross-validation accuracy of kernel-SVM classifiers for \gls{BA} model
    parameters, with kernel meta-parameters $\lambda = 0.3$ and $\sigma = 4$.
    Each point was calculated based on 300 simulated transmission trees over
    networks with three different values of the parameter being tested.
    Vertical lines are empirical 95\% confidence intervals based on 1000 2-fold
    cross-validations.
  }
  \label{fig:rsquared}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{kernel-alpha-crossv}
  \caption{
    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
    of \gls{BA} network model, for various tree kernel meta-parameters and
    epidemic scenarios. Each point was calculated based on 300 simulated
    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
    and and dashed lines indicate, respectively, performance of SVM using the
    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
    are number of infected nodes before the simulation was stopped ($I$) and
    number of tips in the sampled transmission tree.
  }
  \label{fig:crossv}
\end{figure}

@@

Of the four parameters, $\alpha$ was the most accurately estimated, with a
median [IQR] absolute error of 
    0.11 
    [0.05-0.18].
(\cref{fig:abcpt,tab:abchpd}). The accuracy of the estimates was not
significantly different between values of $m$ or $I$ (both one-way ANOVA,
    $p$ = 0.1
and 
    0.25),
although the errors when the true value of $\alpha$ was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
The error in the estimated value of $I$ was
    305.66 
    [107.76-606.59].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
and for $I$ = 2000
    ($p$ = \ensuremath{1.58\times 10^{-6}}),
but not for any values of $m$
    (one-way ANOVA, $p$ = 0.33).
The $m$ parameter was estimated correctly in
    37 \%
of simulations, with an error of one in
    40 \%
and of two or more in 
    22 \%
(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
and $I$ did not significantly affect the error
    (one-way ANOVA, $p$ = 0.5 and
                          0.68).
%but the accuracy was significantly lower for integral than non-integral
%values of $\alpha$
%    (Wilcoxon rank-sum test, $p$ = alpha.m.test$p.value).
Finally, the total number of nodes $N$ was consistently over-estimated by about
a factor of two
    (error \ensuremath{6.59\times 10^{3}} 
    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
No other parameters influenced the accuracy of the $N$ estimates 
    (one-way ANOVA, $p \geq$ 0.72).

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{abc-point-estimate}
  \caption{
    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
    on simulated phylogenies. Dotted lines indicate the true values, and limits
    of the $y$-axes are the regions of uniform prior density.
  }
  \label{fig:abcpt}
\end{figure}

\begin{table}[ht]
  \centering
  \input{\tablepath/abc-hpd}
  \caption{Average \gls{HPD} interval widths for \gls{ABC} model parameter
           estimates.}
  \label{tab:abchpd}
\end{table}

The dispersion of the \gls{ABC} approximation to the posterior also varied
between the parameters, with narrower \gls{HPD} intervals for those with the
most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
distributions for for one simulation (equivalent plots for all the simulations
can be found in the supplemental materials). \gls{HPD} intervals around
$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
whereas the intervals for $m$ and $N$ were more widely dispersed.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\figwidth]{abc-posterior-example}
  \caption{Marginal estimated posterior distributions obtained with ABC on a
  single simulated dataset. Dotted lines and shaded polygon indicate true
  values.}
  \label{fig:abcex}
\end{figure}

To test the effect of model misspecification, we simulated one network where
the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
HPD] estimates for each parameter were: 
$\alpha$, 
  1 [0.85-
                           1.12];
$I$,
  5.23 [2.05-
                       5.85];
$m$,
  \ensuremath{10^{4}} [\ensuremath{3.06\times 10^{3}}-
                              \ensuremath{1.5\times 10^{4}}];
$N$,
  \ensuremath{1.17\times 10^{3}} [505.71-
                       \ensuremath{1.75\times 10^{3}}].

To test the effect of sampling bias, we sampled one transmission tree in a
peer-driven fashion, where the probability to sample a node was twice as high
if one of it's peers had already been sampled. The parameters for this
experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
estimated values were
$\alpha$, 
  0.48 [0.03-
                           0.83];
$I$,
  3.6 [2.18-
                       5.9];
$m$,
  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
                              \ensuremath{1.48\times 10^{4}}];
$N$,
  \ensuremath{2.48\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
                       \ensuremath{3.87\times 10^{3}}].

@@

Contact networks can have a strong influence on epidemic progression, and are
potentially useful as a public health tool. Despite this, few methods exist for
investigating contact network parameters in a phylodynamic framework.
Kernel-ABC is a model-agnostic method which can be used to investigate any
quantity that affects tree shape. In this work, we developed a kernel-ABC-based
method to infer the parameters of a contact network model. The method is
general, and could be applied to any model from which contact networks can be
simulated. We demonstrated the method on the \gls{BA} model, which is a simple 
model incorprorating the preferential attachment feature commonly seen in
real-world networks.
